Model Walkthru for Peloquin & Frank (2016)

### 2017-10-20 paper01_peloquin_frank2016/model_walkthru/paper01_peloquin_frank2016.rmd
push pre-final version in case anyone is working on this before the meeting

---
title: "Determining the alternatives for scalar implicature"
subtitle: "Peloquin & Frank (2016)"
author: "timothy leffel"
date: "oct20/2017"
---

```{r}
### ADAPTED CODE FROM PACKAGE VIGNETTE FOR DEMO OF fullRecursion()
source("setup.r")

@@

```{r}
# source("setup.r")

@@

str(pf_data)
# pf_data %>% filter(words=="none") %>% View() # ***

# carve out the <none, some, all> scale (both expts 1 and 2) 
# pf_data %>% filter(scale=="some_all") %>% View()
pf3_NSA <- pf_data %>% 
  filter(dataset=="pf3") %>% 
  filter(scale=="some_all") %>% 
  select(-exp, -scale, -dataset) %>% 
  select(stars, words, study1, study2, speaker.p) %>% 
  arrange(stars, words)

@@

@@

```{r eval=FALSE, include=FALSE}
lefftpack::lazy_setup()
# devtools::install_github("benpeloquin7/rrrsa")
library("rrrsa")

### CONTENTS -------

# SIX DATASETS

# THREE HELPER FUNCTIONS

# SIX ANALYSIS FUNCTIONS

### DATASETS ------------------------------------------------------------------

# note: on any dataset `df`, start by calling `?df` to see info

# example docs for `rrrsa::peloquinFrank_2Alts`
# 
#   This data set contains empirical literal listener semantics and pragmatic
#   judgments for entailment items (two alternatives) from Peloquin & Frank 2016
# 
#   1) quantityVarName :: "stars"
#   2) semanticsVarName :: "speaker.p"
#   3) itemVarName :: "words"
#   4) groupVarName :: "scale"
#   other) experiment number :: "exp"
#   other) Pragmatic judgments study 1 (e6), study 2 (e11)
# 
#   see also: https://github.com/langcog/scalar_implicature

# 136x16 df, data from frank et al. (under review)
rrrsa::d_pragmods

# 50x7 df, data from pel+frk16, 
rrrsa::peloquinFrank_2Alts

# 75x7 df
rrrsa::peloquinFrank_3Alts

# 100x7 df
rrrsa::peloquinFrank_4Alts

# 125x7 df 
rrrsa::peloquinFrank_5Alts

### HELPER FUNCTIONS ----------------------------------------------------------

# converts vec2 to whatever type vec1 is
rrrsa::rsa.convertVecType(vec1, vec2)

# this is just `function(x) x/sum(x)` for positive `x`
rrrsa::rsa.normVec(v)

# just changes desired column names 
rrrsa::rsa.renameCol(df, currNames, replacements)

### ANALYSIS FUNCTIONS --------------------------------------------------------

### `rrrsa::rsa.informativity(m_u, alpha, cost)` -------
# ifelse(m_u == 0, 0, exp(-alpha * (-log(m_u) - cost)))
rrrsa::rsa.informativity(m_u, alpha, cost)

### `rrrsa::rsa.utility(items, costs, alpha)` ------
# rsa.normVec(mapply(rsa.informativity, items, costs, alpha = alpha))
rrrsa::rsa.utility(items, costs, alpha)

### `rrrsa::rsa.reason(m, costs, priors, depth, alpha)` ----
rrrsa::rsa.reason(m, costs, priors, depth, alpha, usePriorEveryRecurse)

### `rrrsa::rsa.runDf()` ----------------
rrrsa::rsa.runDf(data, quantityVarName, semanticsVarName, 
                 itemVarName, costsVarName, priorsVarName, depth, alpha,
                 usePriorEveryRecurse)

### `rrrsa::rsa.tuneDepthAlpha()` --------
rrrsa::rsa.tuneDepthAlpha()

### rrrsa::rsa.fullRecursion(m, costs, priors, alpha) ------
# `m` a matrix
# `costs` a vector of costs, w length == ncol(m) (defaults to zeros)
# `priors` vector of priors, w length == nrow(m) (defaults to ones)
# `alpha` scalar noise param, defaults to 1 
# returns a posterior distribution 
rrrsa::rsa.fullRecursion(m, costs, priors, alpha)

>>>>

---
title: "Model Walkthru for Peloquin & Frank (2016)"
subtitle: "\"Determining the alternatives for scalar implicature\""
author: "timothy leffel, oct20/2017"
output: 
  html_document:
    css: style.css
    self_contained: no
---

```{r setup}
source("setup.r")
```

## Quick recap of paper

## Contents of model implementation package

Below is a listing of the contents of the `rrrsa::` package.

#### Five datasets 
```{r}
# FIVE DATASETS
rsa_data <- list(
  # from frank et al submitted:
  frank = rrrsa::d_pragmods,
  # from pelfrank16: 
  pf2   = rrrsa::peloquinFrank_2Alts,
  pf3   = rrrsa::peloquinFrank_3Alts,
  pf4   = rrrsa::peloquinFrank_4Alts,
  pf5   = rrrsa::peloquinFrank_5Alts
)
```

#### Three helper functions 
```{r}
# THREE HELPER FUNCTIONS
rsa_helper <- list(
  # converts vec2 to whatever type vec1 is
  convertVecType = rrrsa::rsa.convertVecType, 
  # this is just `function(x) x/sum(x)` for positive `x`
  normVec        = rrrsa::rsa.normVec, 
  # just changes desired column names 
  renameCol      = rrrsa::rsa.renameCol
)
```

#### Six analysis functions 
```{r}
# SIX ANALYSIS FUNCTIONS
rsa_model <- list(
  # computes informativity given params `m_u`, `alpha`, `cost`
  informativity  = rrrsa::rsa.informativity,
  # computes utility given params `items`, `costs`, `alpha`
  utility        = rrrsa::rsa.utility,
  # wrapper that runs `depth`-many iterations of `rsa.fullRecursion()` 
  reason         = rrrsa::rsa.reason,
  # wrapper around `rsa.reason` that accepts input as a data frame 
  runDf          = rrrsa::rsa.runDf,
  # explores correlation between data and model preds for varying `alpha` 
  tuneDepthAlpha = rrrsa::rsa.tuneDepthAlpha,
  # main function implementing RSA model 
  fullRecursion  = rrrsa::rsa.fullRecursion
)
```

## An example in action

```{r}
### ADAPTED CODE FROM PACKAGE VIGNETTE FOR DEMO OF fullRecursion()

@@

## Reproducing the paper results

#### get acquainted with the datasets, combine them

```{r}
source("setup.r")

@@

str(pf_data)
# pf_data %>% filter(words=="none") %>% View() # *** 
```

#### Specific example: Expt1, three alts, <*none, some, all*>

```{r}
# carve out the <none, some, all> scale (both expts 1 and 2) 
# pf_data %>% filter(scale=="some_all") %>% View()
pf3_NSA <- pf_data %>% 
  filter(dataset=="pf3") %>% 
  filter(scale=="some_all") %>% 
  select(-exp, -scale, -dataset) %>% 
  select(stars, words, study1, study2, speaker.p) %>% 
  arrange(stars, words)

@@

#### Specific example: Expt2, three alts, <*none, some, all*>

#### Specific example: ...

## Play around with parameters

@@

## Concepts used in RSA models

#### primitive principles

- speech-acts are rational behaviors -- actions in **signaling games** 
- utterances are selected by their **expected utility** 
- the utility of an utterance is a function of its **informativity** and its **cost**

#### model of a conversational exchange

- a **speaker** $S_n$ wants to communicate some proposition 
- $S_n$ selects utterance $u$ from a set of alternative utterances 
- an actual ("pragmatic") **listener** ???
- FINISH

## The mathematics underlying RSA models

The following equations are from Goodman & Frank's (2012) model specification.

#### equation 1 (GF2012)

$$P(r_s|w, C) = \frac{P(w|r_s, C)\times P(r_s)}{\sum\limits_{r'\in C} P(w|r', C) \times P(r')}$$

where

- $r_s$ is an intended referent (element of $C$), 
- $C$ is a context (set of objects $r, r', ...\in C$), and 
- $w$ is the word uttered by the speaker. 
- The **prior** is $P(r_s)$, 
- the **likelihood** is $P(w|r_s, C)$, and 
- the **normalizing constant** is $\sum\limits_{r'\in C} P(w|r', C) \times P(r')$

#### equation 2 (GF2012)

$$P(w|r_s, C) = \frac{|w|^{-1}}{\sum\limits_{w'\in W} |w'|^{-1}}$$

where 
- $|w|$ is the number of objects in $C$ to which word $w$ could(?) apply, 
- and $W$ is the set of words that apply to the intended referent $r_s$.

##### equation S1 (GF12 supplement)

$$P(w|r_s, C) \propto e^{\alpha\times U(w; r_s, C)}$$

where

- the **context** $C = \{o_1, ..., o_n\}$ is a set of objects 
- the **vocabulary** is a set of words $V = \{w_1, ..., w_m\}$ 
- each word $w\in V$ has a meaning $[\![ w ]\!] : C \rightarrow \{0, 1\}$ 
- words are chosen in proportion to their expected utility

#### equation S2 -- utility (GF12 supplement)

$$U(w; r_s, C) = I(w; r_s, C) - D(w)$$

where

- $I(w; r_s, C)$ is the informativeness of utterance $w$ w.r.t. intended referent $r_s \in C$, and 
- $D(w)$ is $w$'s cost (in $C$?). (for single words, $D : V \rightarrow \mathbb{R}^+$ is constant)

##### surprisal (GF12 supplement)

for a sample $x$ from a known distribution $p(x)$, the surprisal of $x$ is
$$I_p(x) = -log(p(x))$$

##### utility and surprisal (GF12 supplement)

utility decreases with surprisal:

$$I(w; r_s, C) = -I_{\widetilde{w}_C}(r_s)$$

where $\widetilde{w}_C$ is the distribution over objects that would come froma  aliteral interpretation of $w$ in context $C$.

##### equation S3 -- literal listener model derivation (GF12, supplementary materials)

"if listeners interpret the utterance $w$ literally, assigning zero probability to objects for which the word is false, they assign equal probability to each object conssitent with $w$. This distribution over objects can be written:"

$$\widetilde{w}_C(o) = 
    \begin{cases}
    \frac{1}{|w|} & \text{if } w(o) = true\\\\
    0             & \text{otherwise }
    \end{cases}
$$

##### equation S4 -- listener likelihood (GF12, supplement)

equation S4 follows from S1-S3, which is equivalent to equation 2 (aka "the size principle")

$$P(w|r_s, C) = \frac
                    {e^{-(-\log(|w|^{-1}))}}
                    {\sum\limits_{w'\in V \text{s.t. } w'(r_s) = true} e^{-(-\log(|w'|^{-1}))}}$$

"Thus in our experiments, the speaker's abstract goal of being informative reduces to a simple formulation: whoose a word that applies to the referent and picks out a relatively smaller section of the context. Listeners may then use this model of a speaker as their likelihood function, to be combined with prior information about contextual salience as in Equation 1 in the main text."

<!-- A `pragmatic listener` $P_{L_n}(m|u)$, reasons about intended meaning $m$ of an utterance $u$ by a `rational speaker` $P_{s_n}(u|m)$ who chooses an utterance according to the expected utility of an utterance $U(m;u)$. $\alpha$ is a decision noise parameter. -->

<!-- $$P_{L_n}(m|u) \propto P_{S_n}(u|m)P(m)$$ -->
<!-- $$P_{S_n} \propto e^{U(m;u)}$$ -->
<!-- $$U(m;u) = -\alpha(-\log(P_{L_{n-1}}(m|u)) - C(u))$$ -->

<!-- See Frank & Goodman (2012) and Goodman & Stuhmuller (2013) for the original descriptions of the RSA framework. Frank, et al. (Under Review) also provides a comprehensive presentation and evaluation of RSA. -->

```{r eval=FALSE, include=FALSE}
### rrrsa::rsa.fullRecursion(m, costs, priors, alpha) ------
# `m` a matrix
# `costs` a vector of costs, w length == ncol(m) (defaults to zeros)
# `priors` vector of priors, w length == nrow(m) (defaults to ones)
# `alpha` scalar noise param, defaults to 1 
# returns a posterior distribution 
rrrsa::rsa.fullRecursion(m, costs, priors, alpha)
